Distributed Topology Discovery in Self-Assembled Nano Networks

### 2014-02-14 trunk/PAPER/disr.tex
Rearranged, fixed

\section{Detailed Description of \disr{}}
\subsection{Message Types Required}

@@

\subsection{\disr{} Data Structures}
\label{ssec:disr_dstruct}

We distinct between two different kind of data stored at each node:
\emph{Local Environment Data} (\emph{LED}) and \emph{Dynamic Behaviour
Status} (\emph{DBS}).

The \emph{LED} is like a snapshot of the \disr{} algorithm at each node,
consisting in the following variables:
\begin{itemize}
\item{\emph{segID}}:a value used to specify the segment to which the
node has been assigned or is candidate for being assigned.
\item{\emph{visited}}: a boolean value. When \emph{true}
a \emph{segID} different from \texttt{NULL} specifies the segment 
to which the node has been assigned. 
% TODO: the following is true, but complicates things at this stage
%Note that a value different from \emph{NULL} does not
%necessarily mean the node has been assigned to some segment (e.g. when
%searching starting segment is set \emph{visited} by default even if not yet
%assigned to any segment). 
\item{\emph{tvisited}}: a boolean value. If \emph{true}, the node is
being considered as candidate for a segment, and the \emph{segID} value
specifies the segment id for which the node is candidate. 
\item{\emph{link\_visited[]}}: an array
of values representing information about attached links, that is, the
segID of the segment owning each link. When \texttt{NULL}, the corresponding link has not yet been
assigned.
%TDB: do we need to introduce some value for the
%“bridge” status ? (e.g. a node connecting the terminal node of a
%subnet with the starting one of another) 
\item{\emph{link\_tvisited[]}}: an array of
values representing information about attached links, that is, the \emph{segID} of
the segment for which the link is candidate. When \texttt{NULL}, the link is
currently not \emph{tvisited}.  
\end{itemize}
%\item{\emph{starting}}: boolean, \emph{true} if the node is a
%starting node from which the whole process was initiated.
%\end{itemize}

In addition to the \emph{LED} variables described above, further
information should be stored in order to capture the current dynamic
behaviour of the node. This is represented by \emph{DBS} variable,
which strictly depends on the \emph{LED} data and the events occurring
at the node. The \emph{DBS} can have the following values:

\begin{itemize}
\item{\emph{Free}}: a node that has  not been yet considered  by the \disr{} algorithm.
\item{\emph{Bootstrap}}: a node which has been explicitly set as bootstrap node from
an upper layer via. 
\item{\emph{ActiveSearching}}: a node from which a find process of new segments has
been started and not yet cancelled or confirmed. 
%This happens when all
%the following conditions are matched: the node is marked as \emph{visited},
%that is \emph{visited} = TRUE 
%TODO: \emph{segID} is set to some value reflecting
%the segment request sent there is a link that is being investigated by
%the node, that is a tvisited\_link marked with the same id of the node 
\item{\emph{Candidate}}: a node currently candidate for belonging to some segment
, but not being itself the node from which the find process was started.

%The following conditions should be matched: \emph{tvisited} =
%true, with segID=X different from node's id.  Two links set as
%\emph{link\_tvisited} with id X. Let’s refer them as  links 'i' and 'j'. Then
%link\_tvisited[i] was set = X when the find process from an adjacent
%node reached the current node, while link\_tvisited[j] has been set to
%X when the node itself started to investigate its free links for the
%segment request with id X. Note that while investigating its attached
%links, if the find process fails along path 'j', the current node can
%investigate another free node 'k' if suitable.  
\item{\emph{CandidateStarting}}: same as above, but the node is currently being considered as candidate
for starting segment. 
%Two main differences: Can confirm
%\texttt{STARTING\_SEGMENT\_CONFIRM} with the same \emph{segID} when receiving a normal
%\texttt{SEGMENT\_REQUEST}: Overwrites its status setting it to \emph{Candidate} for the
%segID frees the links previously set as \emph{tvisited} during flooding
\item{\emph{Assigned}}: a node for which the segment has been determined.  
%The \emph{visited} value is \emph{true} NOTE: an \emph{Assigned} state is a quick temporary
%state, since it becomes \emph{ActiveSearching} if a not (\emph{visited}/tvisited)
%link is suitable for searching new segments (see next\_visited
%procedure on paper) 
\end{itemize}

A quantitative analysis of the resources needed to implement these
fields is presented in Section~\ref{sec:implementation} when
discussing the impact of the \disr{} control logic and storage on
hardware implementation of the node.

@@

\item{\textbf{Failing while searching a segment}}: this happen when a node received a \texttt{SEGMENT\_REQUEST} packet but
matches one of two the following conditions: first, the node is $Free$ but has no
more suitable free links (thus can’t forward the
\texttt{SEGMENT\_REQUEST})$(f)$; the node
is already \emph{Candidate} with another segment id, i.e. another find
process.  In all these cases the node
sends back a \texttt{SEGMENT\_CANCEL} along the incoming link (see at
the bottom of Figure~\ref{fig:disr_events}).
%Each of the subsequent nodes will forward this \texttt{SEGMENT\_CANCEL}
%only if no more free links can be explored starting from them.
\end{itemize}

% TODO: FIX and RE-enable in long paper

@@

The \emph{LED} variables at each node are are initially set so the node and
each of its links is set as free. The only node that makes exception
is the bootstrap node, initialized in the \emph{Bootstrap} status. All
the subsequent events happening at each node are the consequence of
its current dynamic status (DBS) and the packets received.

%TODO: add state machines

% TODO: fix, move elsewhere or delete
%Assumption: when discussing of visited/\emph{tvisited} nodes, we are assuming
%with an id corresponding to the same subnet. Nodes involved in other
%subnet processes can simply ignore these packets. (TBD: formalize more
%clearly in which cases inter-subnet packets must be ignored, probably
%this happens most of the times, but in some cases, e.g. when a
%terminal node is found, some kind of communication packets are needed
%in order to crearte the bridge between two subnets.)

@@

The request for the first segment should be managed differently since
all nodes (except the starting one) must forward the packet using a
flooding mechanism. This is necessary since the request packet must
return the bootstrap node.
When a node receives a \texttt{STARTING\_SEGMENT\_REQUEST}, the
following cases can happen:

\begin{itemize}
\item The node is not candidate or assigned: it should set itself as
\emph{Candidate} and forward the packet along its free links, using
flooding.  
\item The node is \emph{Candidate} with the same
\emph{segID} and \emph{src\_id} is equal to node id: this means the
node was the initiator of the request, so a starting segment was found and
should be confirmed sending back a \texttt{STARTING\_SEGMENT\_CONFIRM}
packet.  
\item The node is \emph{Candidate}, with the same \emph{segID} of the
packet but \emph{src\_id} is different from node id: due
the flooding mechanism, another starting segment packet reached a node
which was previously marked as \emph{Candidate} from the same starting segment
search process. Since the node is candidate with the same id, it means
that it already accomplished in the task of propagating that kind
packets, thus can simply ignore the event thrashing the packet.  
%TBD: how the status should be changed ? depend on the flooding mechanism. Note
%that these links do not include the link from which the request was
%received.  
\item The node is \emph{Candidate}, with a different \emph{segID}: 
this simply means that the starting segment received is deprecated,
because the node has already accepted non-starting segment requests
originated from other \emph{Assigned} nodes.
\item The node is \emph{Assigned}, with a different \emph{segID}:
same as the previous case.

@@

\begin{itemize}
\item The node is \emph{Assigned}: a segment was
found and it should be confirmed sending back a \texttt{SEGMENT\_CONFIRM}
packet.  
\item The node is \emph{Candidate}: it should discard the packet sending
back a \texttt{SEGMENT\_CANCEL} 
\item The node is not \emph{Assigned} or \emph{Candidate} and has free
links: it marks itself as \emph{Candidate} and forwards the \texttt{SEGMENT\_REQUEST}
to one of its free links, according to some internal ordering
\end{itemize}

NOTE: the main difference between confirming a \texttt{STARTING\_SEGMENT\_REQUEST}
and confirming a \texttt{SEGMENT\_REQUEST} is that in the first case the node
itself is included in the segment , while in the second case the
receiving node does not belong to the segment found.

@@

\subsection{Receiving a \texttt{SEGMENT\_CONFIRM}}
When received, if the node status is \emph{Candidate} with a \emph{segID} corresponding to the one indicated in
the packet, the node learns to belong to the segment associated to the
id. Further, it should forward this packet to the link where the
original \texttt{SEGMENT\_REQUEST} packet came from, so that all candidate nodes
can learn the segment id they belong to.  \emph{LED} shoud be
updated from \emph{tvisited} to \emph{visited}.  the
\emph{link\_tvisited} previously set with the \emph{segID} should be invalidated the
\emph{link\_visited} variable associated to the incoming confirm should be set
as segID. Note that if the request was flooded across different links,
the \emph{link\_visited} variable associated to the other paths should not be
set.  the \emph{link\_visited} variable associated to the link from which the
request was originated should be set segID

\subsection{Receiving a \texttt{SEGMENT\_CANCEL}}
When a node receives a \texttt{SEGMENT\_CANCEL} packet it means that searching
for a segment along that path was unsuccessful. But if the node still
has other free links to try, it should forward a \texttt{SEGMENT\_REQUEST} to
the next link (next according to some internal ordering). So a node
forwards back the \texttt{SEGMENT\_CANCEL} packet along the link that originated
the \texttt{SEGMENT\_REQUEST} packet only when there are no more free links to
try. If this is the case, the node modifies its status from
\emph{Candidate} to \emph{Free} and forwards the \texttt{SEGMENT\_CANCEL} packet to the link from which
the request was received. The process stops when the \texttt{SEGMENT\_CANCEL}
packet reach the starting node that originated the request.

%TODO REPEAT!!
%When a node discovers that the path that includes
%the node itself is no more suitable, a \texttt{SEGMENT\_CANCEL} is sent back
%across the link from which the \texttt{SEGMENT\_REQUEST} was received.  Possible
%reasons for that: The node it’s already candidate for another segment
%The node it’s free but can’t find a free link (according the
%cyclelinks timeout) The node it’s \emph{tvisited} for the same request id,
%previously forwarded that request along some free direction, but now
%it’s receiving a a cancel request and has no more free links
\subsection{Intra-node vs Inter-node parallelism}
In the processes described above, we assumed that each node in the
\emph{Assigned} status can start a find segment process by injecting a
\texttt{SEGMENT\_REQUEST}. However, a more accurate decision when
defining \disr{} should be made in order to decide whether or not the
node must actually perform this action.  The point above is is
strictly related to the general question: which level of “parallelism”
should we allow in \disr{} ? The adopted approach is that, although
the nature of \disr{} itself is intrinsicly parallel, the use of
parallelism makes things work in a more complex way. In other words,
\disr{} is parallel when needed, but it does not exploit parallelism
as an “improving feature”. When not needed, things should be
serialized. For example: when a node is “ready searching” could start
several find segment process associated to the same segment, one for
each free link. But serializing this by investigating the free links
in order could be a simplest solution. We refer to this saying that we
avoid intra-node parallelism.  Note that while intra-node parallelism
can be avoided, trying to avoid inter-node parallelism could be more
complex than allowing the parallelism itself. Imagine for example the
effort of trying to coordinate nodes so that an unique finding node
process is actually running in the whole subnet. Thus, in contrast to
the intra-node parallelism, the inter-node parallelism is a structural
property of the \disr{} algorithm and should not be avoided.
%------------------------------------------------------------------------------

>>>>

\subsection{\disr{} Data Structures}
\label{ssec:disr_data}

We distinct between two different kind of data stored at each node:
\emph{Dynamic Behaviour Status} (\emph{DBS}) and \emph{Local
Environment Data} (\emph{LED}).

The \emph{DBS} stores information in order to capture the dynamic
behaviour of the node  in the context of the whole
\disr{} execution phased. The \emph{DBS} can have the following values:

\begin{itemize}
\item{\emph{Free}}: a node that has  not been yet considered  by the \disr{} algorithm.
\item{\emph{Bootstrap}}: a node which has been explicitly set as bootstrap node from
an upper layer via. 
\item{\emph{ActiveSearching}}: a node from which a find process of new segments has
been started and not yet cancelled or confirmed. 
\item{\emph{Candidate}}: a node currently candidate for belonging to some segment
, but not being itself the node from which the find process was started. 
\item{\emph{CandidateStarting}}: same as above, but the node is currently being considered as candidate
for starting segment. 
\item{\emph{Assigned}}: a node for which the segment has been determined.  
\end{itemize}

The \emph{LED} are like a instanst snapshot of the \disr{} algorithm at each node,
consisting in the following variables:
\begin{itemize}
\item{\emph{segID}}:a value used to specify the segment to which the
node has been assigned or is candidate for being assigned.
\item{\emph{visited}}: a boolean value. When \emph{true}
then a \emph{segID} different from \texttt{NULL} specifies the segment 
to which the node has been assigned. 
\item{\emph{tvisited}}: a boolean value. If \emph{true}, the node is
being considered as candidate for a segment, and the \emph{segID} value
specifies the segment id for which the node is candidate. 
\item{\emph{link\_visited[]}}: an array
of values representing information about attached links, that is, the
segID of the segment owning each link. When \texttt{NULL}, the corresponding link has not yet been
assigned.
\item{\emph{link\_tvisited[]}}: an array of
values representing information about attached links, that is, the \emph{segID} of
the segment for which the link is candidate. When \texttt{NULL}, the link is
not candidate for any segment, i.e. maybe already assigned
or free.
\end{itemize}

\subsection{Message Types Required}

@@

A quantitative analysis of the resources needed to implement and
manage these structures is presented in Section~\ref{sec:implementation} when
discussing the impact of the \disr{} control logic and storage on
hardware implementation of the node.

@@

\item{\textbf{Failing while searching a segment}}: this happen when a node received a \texttt{SEGMENT\_REQUEST} packet but
matches one of two the following conditions: first, the node is $Free$ but has no
more suitable free links (thus can’t forward the
\texttt{SEGMENT\_REQUEST})$(f)$; the node
is already \emph{Candidate} with another segment id, i.e. another find
process.  In all these cases the node
sends back a \texttt{SEGMENT\_CANCEL} along the incoming link (see at
the bottom of Figure~\ref{fig:disr_events}).
\end{itemize}

@@

In this Section we provide further details about \disr{}
execution model, describing the internal behavior implemented at
each node. Note that this also corresponds exactly to what should be
implemented from the node-hardware perspective, as discussed in
Section~\ref{sec:implementation}.

The \emph{LED} variables at each node
($segID$,$visited$,$tvisited$,\emph{link\_visited[]},\emph{link\_tvisited[]})
are initially set as \texttt{NULL}, so the node and each of its links
are considered as free.
The only node that makes exception is the bootstrap node, initialized
in the \emph{Bootstrap} status. All the subsequent events happening at
each node are the consequence of its current dynamic status (DBS), the
packets received and the internal status represented by the
\emph{LED}. Note that in the following, when using expression like
``has a free link'' or ``free node'' we will assume the relationship beetween 
$LED$ and $DBS$ discussed in Section~\ref{sec:disr_data}. Further, for
each packet type all the cases not explicitly listed can simply be
ignored since are associated to invalid/inconsistent events of the
\disr{} execution model. A complete list of these cases can be checked
in the environment simulator source code at~\ref{nanoxim}.

@@

The request for the first segment should be managed differently since
all nodes (except the bootstrap one) must forward the packet using a
broadcasting mechanism. This is necessary since the request packet must
return the bootstrap node.
When a node receives a \texttt{STARTING\_SEGMENT\_REQUEST}, the
following cases can happen:

\begin{itemize}
\item The node is $Free$: it should set itself as
\emph{CandidateStarting} and forward the packet along its free links, using
broadcasting and marking these links as $tvisited$ with $segID$ found
in the packet.
\item The node is \emph{CandidateStarting} with the same
\emph{segID} and \emph{src\_id} of the packet is equal to node id:
this means that the node was the initiator of the request, thus a starting segment has
just been found and should be confirmed sending back a \texttt{STARTING\_SEGMENT\_CONFIRM}
packet.  
\item The node is \emph{CandidateStarting}, with the same \emph{segID} of the
packet but the \emph{src\_id} field is different from node id: since the node is candidate with the same id, it means
that it already accomplished the task of propagating that kind of
packets, thus can simply ignore the event thrashing the packet.  
\item The node is \emph{Candidate}, with a different \emph{segID}: 
this simply means that the starting segment request just received is deprecated,
because the node has already accepted a non-starting segment request
that originated from another \emph{Assigned} nodes.
\item The node is \emph{Assigned}, with a different \emph{segID}:
same as the previous case.

@@

\begin{itemize}
\item The node is \emph{Assigned}: a segment should be confirmed
sending back a \texttt{SEGMENT\_CONFIRM} packet.
\item The node is \emph{Candidate}: it should discard the packet sending
back a \texttt{SEGMENT\_CANCEL} 
\item The node is \emph{Free} and has free
links: it marks itself as \emph{Candidate} and forwards the \texttt{SEGMENT\_REQUEST}
to one of its free links, according to some internal ordering.
\item The node is \emph{CandidateStarting}: same as being $Free$,
since a segment request circulating in the network indicates that the
starting segment process has been completed.
\end{itemize}

Note that the main difference between confirming a \texttt{STARTING\_SEGMENT\_REQUEST}
and confirming a \texttt{SEGMENT\_REQUEST} is that in the first case the node
itself is included in the segment.

@@

\subsection{Receiving a \texttt{SEGMENT\_CONFIRM}}
When the node status is \emph{Candidate} with a \emph{segID} corresponding to the one indicated in
the packet: the node set itself as $Assigned$ to the segment $SegID$. Further, it should forward this packet to the link where the
original \texttt{SEGMENT\_REQUEST} packet came from, so that all candidate nodes
can learn the segment id they belong to. \emph{LED} shoud be
updated from \emph{tvisited} to \emph{visited}.

\subsection{Receiving a \texttt{SEGMENT\_CANCEL}}
When a node receives a \texttt{SEGMENT\_CANCEL} packet it means that
searching for a segment along that path was unsuccessful. But if the
node still has some other free link to try, it should forward a
\texttt{SEGMENT\_REQUEST} to the that link. So a node forwards back
the \texttt{SEGMENT\_CANCEL} packet along the link that originated the
\texttt{SEGMENT\_REQUEST} packet $only$ when there are no more free
links to try. If this is the case, the node modifies its status from
\emph{Candidate} to \emph{Free} and forwards the
\texttt{SEGMENT\_CANCEL} packet to the link from which the request was
received. The process stops when the \texttt{SEGMENT\_CANCEL} packet
reach the starting node that originated the request.

\subsection{Intra-node vs Inter-node parallelism}
In the processes described above, we assumed that each node in the
\emph{Assigned} status can start a find segment process by injecting a
\texttt{SEGMENT\_REQUEST}. However, a more accurate decision when
defining \disr{} should be made in order to decide whether or not the
node must actually perform this action.  The point above is is
strictly related to the general question: which level of “parallelism”
should we allow in \disr{} ? The adopted approach is that, although
the nature of \disr{} itself is intrinsicly parallel, the use of
parallelism makes things work in a more complex way. In other words,
\disr{} is parallel when needed, but it does not exploit parallelism
as an “improving feature”. When not needed, things should be
serialized. For example: when a node is “ready searching” could start
several find segment process associated to the same segment, one for
each free link. But serializing this by investigating the free links
in order could be a simplest solution. We refer to this saying that we
avoid intra-node parallelism.  Note that while intra-node parallelism
can be avoided, trying to avoid inter-node parallelism could be more
complex than allowing the parallelism itself. Imagine for example the
effort of trying to coordinate nodes so that an unique finding node
process is actually running in the whole subnet. Thus, in contrast to
the intra-node parallelism, the inter-node parallelism is a structural
property of the \disr{} algorithm and should not be avoided.
%------------------------------------------------------------------------------
